---
title: "Assignment 5 Processing Data"
author: "Alex Martinez"
editor: visual
---

### My Data

The data set I chosed is ORal health infromation from children in school. Some important variables it has is

-   **Caries Experience**. Percentage of children with caries experience, including treated and untreated tooth decay.

-   **Untreated Tooth Decay**. Percentage of children with untreated tooth decay.

-   **Dental Sealants**. Percentage of children with dental sealants on at least one permanent molar tooth.

Here is more information from the website below:

Data for School year-end 1994 through year-end 2020. State oral health surveys are the data sources for these indicators. States periodically conduct independent screening surveys of a probability sample designed to be representative of all third-grade students in the state. Some states also conduct surveys of students in other grades in school, or of Head Start program enrollees. This surveillance activity is voluntary. States submit their data to the Association of State and Territorial Dental Directors (ASTDD), where the survey design and data collected are reviewed for quality and against the criteria for inclusion in NOHSS, before being sent to CDC for inclusion in Oral Health Data. For more information,see: <http://www.cdc.gov/oralhealthdata/overview/childIndicators/>

### Exploring my dataset

```{r}
library(tidyverse)
getwd()
data <- read.csv("C:/Users/alexm/Downloads/Pract 2/AlexMartinez-P2-portfolio/cdcdata-exercise/CDCdata.csv")

str(data)
```
First I will select the variables I will work with before I clean my data

```{r}
data <- data %>% select(SchoolYearStart,LocationAbbr, Grade, Indicator,Data_Value, Data_Value_Footnote, Sample_Size, Response_Rate)
str(new)
summary(new)
head(new)

# Convert necessary columns to appropriate data types
data <- data %>%
  mutate(
    Data_Value = as.numeric(Data_Value),
    Sample_Size = as.numeric(Sample_Size),
    Response_Rate = as.numeric(Response_Rate)
  )

str(data)

# Exploratory data analysis
summary_stats <- data %>%
  summarise(
    Data_Value_mean = mean(Data_Value, na.rm = TRUE),
    Data_Value_sd = sd(Data_Value, na.rm = TRUE),
    Data_Value_min = min(Data_Value, na.rm = TRUE),
    Data_Value_max = max(Data_Value, na.rm = TRUE),
    Sample_Size_mean = mean(Sample_Size, na.rm = TRUE),
    Sample_Size_sd = sd(Sample_Size, na.rm = TRUE),
    Sample_Size_min = min(Sample_Size, na.rm = TRUE),
    Sample_Size_max = max(Sample_Size, na.rm = TRUE),
    Response_Rate_mean = mean(Response_Rate, na.rm = TRUE),
    Response_Rate_sd = sd(Response_Rate, na.rm = TRUE),
    Response_Rate_min = min(Response_Rate, na.rm = TRUE),
    Response_Rate_max = max(Response_Rate, na.rm = TRUE)
  )

print(summary_stats)

# Distribution of Data_Value
ggplot(data, aes(x = Data_Value)) +
  geom_histogram(binwidth = 2, fill = 'blue', color = 'black') +
  theme_minimal() +
  labs(title = 'Distribution of Data_Value', x = 'Data_Value', y = 'Frequency')

# Distribution of Sample_Size
ggplot(data, aes(x = Sample_Size)) +
  geom_histogram(binwidth = 500, fill = 'green', color = 'black') +
  theme_minimal() +
  labs(title = 'Distribution of Sample_Size', x = 'Sample_Size', y = 'Frequency')

# Distribution of Response_Rate
ggplot(data, aes(x = Response_Rate)) +
  geom_histogram(binwidth = 5, fill = 'red', color = 'black') +
  theme_minimal() +
  labs(title = 'Distribution of Response_Rate', x = 'Response_Rate', y = 'Frequency')

# Percentage of each category in Grade
grade_dist <- data %>%
  group_by(Grade) %>%
  summarise(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)

print(grade_dist)

# Percentage of each category in Indicator
indicator_dist <- data %>%
  group_by(Indicator) %>%
  summarise(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)

print(indicator_dist)

# Percentage of each category in LocationAbbr
location_dist <- data %>%
  group_by(LocationAbbr) %>%
  summarise(count = n()) %>%`
  mutate(percentage = count / sum(count) * 100)

print(location_dist)
```

